---
title: "extra code ontario food prices"
author: "Alice Huang"
date: "03/01/2023"
output: html_document
---

---
title: "Using SARIMA Models to Forecast Ontario Retail Food Prices Based on Jan 2017 - Oct 2022 Data"
author: "Alice Huang"
date: "20/12/2022"
output: pdf_document
---
```{r include=FALSE}
library(tseries)
library(astsa)
```


```{r warning=FALSE, include=FALSE}
library(tidyverse)
data <- read.csv("statscan_foodprices_jan2017_to_oct2022.csv")
# some column values were originally stored as strings for some reason so they were converted to numeric data type
data[,2] <- as.numeric(data[,2])
data[,3] <- as.numeric(data[,3])
data[,4] <- as.numeric(data[,4])
# just_data has each column being the prices for each food item, since R ts() function wants column vectors as inputs
just_data <- t(data[,2:71])
```

```{r include=FALSE}
# converting the dataframe columns to time series objects
beef_stewing <- ts(just_data[,1], frequency = 12, start = c(2017, 1), end = c(2022, 10))
beef_striploin <- ts(just_data[,2], frequency = 12, start = c(2017, 1), end = c(2022, 10))
beef_top_sirloin <- ts(just_data[,3], frequency = 12, start = c(2017, 1), end = c(2022, 10))
beef_rib <- ts(just_data[,4], frequency = 12, start = c(2017, 1), end = c(2022, 10))
beef_ground <- ts(just_data[,5], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pork_loin <- ts(just_data[,6], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pork_rib <- ts(just_data[,7], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pork_shoulder <- ts(just_data[,8], frequency = 12, start = c(2017, 1), end = c(2022, 10))
chicken_whole <- ts(just_data[,9], frequency = 12, start = c(2017, 1), end = c(2022, 10))
chicken_breast <- ts(just_data[,10], frequency = 12, start = c(2017, 1), end = c(2022, 10))
chicken_thigh <- ts(just_data[,11], frequency = 12, start = c(2017, 1), end = c(2022, 10))
chicken_drumstick <- ts(just_data[,12], frequency = 12, start = c(2017, 1), end = c(2022, 10))
bacon <- ts(just_data[,13], frequency = 12, start = c(2017, 1), end = c(2022, 10))
wiener <- ts(just_data[,14], frequency = 12, start = c(2017, 1), end = c(2022, 10))
salmon <- ts(just_data[,15], frequency = 12, start = c(2017, 1), end = c(2022, 10))
shrimp <- ts(just_data[,16], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_salmon <- ts(just_data[,17], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_tuna <- ts(just_data[,18], frequency = 12, start = c(2017, 1), end = c(2022, 10))
meatless_burger <- ts(just_data[,19], frequency = 12, start = c(2017, 1), end = c(2022, 10))
milk_1L <- ts(just_data[,20], frequency = 12, start = c(2017, 1), end = c(2022, 10))
milk_2L <- ts(just_data[,21], frequency = 12, start = c(2017, 1), end = c(2022, 10))
milk_4L <- ts(just_data[,22], frequency = 12, start = c(2017, 1), end = c(2022, 10))
soy_milk <- ts(just_data[,23], frequency = 12, start = c(2017, 1), end = c(2022, 10))
nut_milk <- ts(just_data[,24], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cream <- ts(just_data[,25], frequency = 12, start = c(2017, 1), end = c(2022, 10))
butter <- ts(just_data[,26], frequency = 12, start = c(2017, 1), end = c(2022, 10))
margarine <- ts(just_data[,27], frequency = 12, start = c(2017, 1), end = c(2022, 10))
block_cheese <- ts(just_data[,28], frequency = 12, start = c(2017, 1), end = c(2022, 10))
yogurt <- ts(just_data[,29], frequency = 12, start = c(2017, 1), end = c(2022, 10))
eggs_dozen <- ts(just_data[,30], frequency = 12, start = c(2017, 1), end = c(2022, 10))
apple <- ts(just_data[,31], frequency = 12, start = c(2017, 1), end = c(2022, 10))
oranges_kg <- ts(just_data[,32], frequency = 12, start = c(2017, 1), end = c(2022, 10))
oranges_1.36kg <- ts(just_data[,33], frequency = 12, start = c(2017, 1), end = c(2022, 10))
banana <- ts(just_data[,34], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pear <- ts(just_data[,35], frequency = 12, start = c(2017, 1), end = c(2022, 10))
lemon <- ts(just_data[,36], frequency = 12, start = c(2017, 1), end = c(2022, 10))
lime <- ts(just_data[,37], frequency = 12, start = c(2017, 1), end = c(2022, 10))
grape <- ts(just_data[,38], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cantaloupe <- ts(just_data[,39], frequency = 12, start = c(2017, 1), end = c(2022, 10))
strawberry <- ts(just_data[,40], frequency = 12, start = c(2017, 1), end = c(2022, 10))
avocado <- ts(just_data[,41], frequency = 12, start = c(2017, 1), end = c(2022, 10))
potatoes_4.54kg <- ts(just_data[,42], frequency = 12, start = c(2017, 1), end = c(2022, 10))
potatoes_kg <- ts(just_data[,43], frequency = 12, start = c(2017, 1), end = c(2022, 10))
sweet_potatoes_kg <- ts(just_data[,44], frequency = 12, start = c(2017, 1), end = c(2022, 10))
tomato <- ts(just_data[,45], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cabbage <- ts(just_data[,46], frequency = 12, start = c(2017, 1), end = c(2022, 10))
carrots <- ts(just_data[,47], frequency = 12, start = c(2017, 1), end = c(2022, 10))
onion_kg <- ts(just_data[,48], frequency = 12, start = c(2017, 1), end = c(2022, 10))
onion_1.36kg <- ts(just_data[,49], frequency = 12, start = c(2017, 1), end = c(2022, 10))
celery <- ts(just_data[50], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cucumber <- ts(just_data[,51], frequency = 12, start = c(2017, 1), end = c(2022, 10))
mushroom <- ts(just_data[,52], frequency = 12, start = c(2017, 1), end = c(2022, 10))
iceberg_lettuce <- ts(just_data[,53], frequency = 12, start = c(2017, 1), end = c(2022, 10))
romaine_lettuce <- ts(just_data[,54], frequency = 12, start = c(2017, 1), end = c(2022, 10))
broccoli <- ts(just_data[,55], frequency = 12, start = c(2017, 1), end = c(2022, 10))
peppers <- ts(just_data[,56], frequency = 12, start = c(2017, 1), end = c(2022, 10))
squash <- ts(just_data[,57], frequency = 12, start = c(2017, 1), end = c(2022, 10))
salad_greens <- ts(just_data[,58], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_french_fries <- ts(just_data[,59], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_green_beans <- ts(just_data[,60], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_broccoli <- ts(just_data[,61], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_corn <- ts(just_data[,62], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_mixed_vegetables <- ts(just_data[,63], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_peas <- ts(just_data[,64], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_pizza <- ts(just_data[,65], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_spinach <- ts(just_data[,66], frequency = 12, start = c(2017, 1), end = c(2022, 10))
frozen_strawberries <- ts(just_data[,67], frequency = 12, start = c(2017, 1), end = c(2022, 10))
white_bread <- ts(just_data[,68], frequency = 12, start = c(2017, 1), end = c(2022, 10))
flatbread_and_pita <- ts(just_data[,69], frequency = 12, start = c(2017, 1), end = c(2022, 10))
crackers_and_crisp_breads <- ts(just_data[,70], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cookies_and_sweet_biscuits <- ts(just_data[,71], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pasta_dry_or_fresh <- ts(just_data[,72], frequency = 12, start = c(2017, 1), end = c(2022, 10))
rice_brown <- ts(just_data[,73], frequency = 12, start = c(2017, 1), end = c(2022, 10))
rice_white <- ts(just_data[,74], frequency = 12, start = c(2017, 1), end = c(2022, 10))
cereal <- ts(just_data[,75], frequency = 12, start = c(2017, 1), end = c(2022, 10))
wheat_flour <- ts(just_data[,76], frequency = 12, start = c(2017, 1), end = c(2022, 10))
white_sugar <- ts(just_data[,77], frequency = 12, start = c(2017, 1), end = c(2022, 10))
apple_juice <- ts(just_data[,78], frequency = 12, start = c(2017, 1), end = c(2022, 10))
orange_juice <- ts(just_data[,79], frequency = 12, start = c(2017, 1), end = c(2022, 10))
coffee_roasted_or_ground <- ts(just_data[,80], frequency = 12, start = c(2017, 1), end = c(2022, 10))
tea <- ts(just_data[,81], frequency = 12, start = c(2017, 1), end = c(2022, 10))
ketchup <- ts(just_data[,82], frequency = 12, start = c(2017, 1), end = c(2022, 10))
vegetable_oil <- ts(just_data[,83], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canola_oil <- ts(just_data[,84], frequency = 12, start = c(2017, 1), end = c(2022, 10))
olive_oil <- ts(just_data[,85], frequency = 12, start = c(2017, 1), end = c(2022, 10))
baby_food <- ts(just_data[,86], frequency = 12, start = c(2017, 1), end = c(2022, 10))
infant_formula <- ts(just_data[,87], frequency = 12, start = c(2017, 1), end = c(2022, 10))
peanut_butter <- ts(just_data[,88], frequency = 12, start = c(2017, 1), end = c(2022, 10))
mayonnaise <- ts(just_data[,89], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_baked_beans <- ts(just_data[,90], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_tomatoes <- ts(just_data[,91], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_soup <- ts(just_data[,92], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_beans_and_lentils <- ts(just_data[,93], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_corn <- ts(just_data[,94], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_peach <- ts(just_data[,95], frequency = 12, start = c(2017, 1), end = c(2022, 10))
canned_pear <- ts(just_data[,96], frequency = 12, start = c(2017, 1), end = c(2022, 10))
dried_lentils <- ts(just_data[,97], frequency = 12, start = c(2017, 1), end = c(2022, 10))
dry_beans_and_legumes <- ts(just_data[,98], frequency = 12, start = c(2017, 1), end = c(2022, 10))
tofu <- ts(just_data[,99], frequency = 12, start = c(2017, 1), end = c(2022, 10))
hummus <- ts(just_data[,100], frequency = 12, start = c(2017, 1), end = c(2022, 10))
salsa <- ts(just_data[,101], frequency = 12, start = c(2017, 1), end = c(2022, 10))
pasta_sauce <- ts(just_data[,102], frequency = 12, start = c(2017, 1), end = c(2022, 10))
salad_dressing <- ts(just_data[,103], frequency = 12, start = c(2017, 1), end = c(2022, 10))
almonds <- ts(just_data[,104], frequency = 12, start = c(2017, 1), end = c(2022, 10))
peanuts <- ts(just_data[,105], frequency = 12, start = c(2017, 1), end = c(2022, 10))
sunflower_seeds <- ts(just_data[,106], frequency = 12, start = c(2017, 1), end = c(2022, 10))
deodorant <- ts(just_data[,107], frequency = 12, start = c(2017, 1), end = c(2022, 10))
toothpaste <- ts(just_data[,108], frequency = 12, start = c(2017, 1), end = c(2022, 10))
shampoo <- ts(just_data[,109], frequency = 12, start = c(2017, 1), end = c(2022, 10))
laundry_detergent <- ts(just_data[,110], frequency = 12, start = c(2017, 1), end = c(2022, 10))
```




```{r}
plot.ts(cantaloupe)
lines(lowess(cantaloupe), col = "red")

plot.ts(diff(cantaloupe))
lines(lowess(diff(cantaloupe)), col = "red")
```
```{r}
acf(cantaloupe)
pacf(cantaloupe)
```


# Analysis and Forecasting of strawberry Prices using SARIMA Model

The monthly strawberry prices show an increasing trend over time. The monthly strawberry prices appear to be seasonal, with a period of 12 months. This is consistent with Ontario weather patterns. Generally, strawberries are only in season during the summer each year. During the winter months, when the strawberries are not in season, the prices are higher. During the summer months, when the strawberries are in season, the prices are lower.

```{r}
plot.ts(strawberry)
lines(lowess(strawberry), col="red")
plot(decompose(strawberry))
```

Since the prices show an increasing trend, the data are not stationary. Since SARIMA models assume that the data are stationary, we transform the data so that it is stationary. The strawberry prices appear to be volatile so we take the log of the strawberry prices. Since the trend appears to be linear, we difference the data. The transformed data appears to have a horizontal trend. We use Dickey-Fuller test to test the null hypothesis that the transformed data is not stationary against the alternative hypothesis that the transformed data is stationary. Since the p-value is 0.01 < 0.05, we have more evidence that the transformed data is stationary.

```{r}
strawberry_transformed <- diff(log(strawberry))
plot.ts(strawberry_transformed)
lines(lowess(strawberry_transformed), col="red")
adf.test(strawberry_transformed)
```

In the acf plot of the  transformed strawberry prices, there appears to be some seasonal behaviour with period 12 months. We seem to have a peak at lags 12, and possibly 24.

The pacf plot seems to cut off after lag 10.

```{r}
acf(strawberry_transformed, lag.max = 48)
pacf(strawberry_transformed, lag.max = 48)
```

We consider an $ARIMA(0,1,0) \times (0,0,1)_{12}$ model.

```{r}
strawberry_sarima1 <- arima(log(strawberry), order = c(0,1,0), seasonal = c(0,0,1))
tsdiag(strawberry_sarima1)
```

```{r}
strawberry_sarima1
```

We test if the coefficients in the SARIMA model are significant. We test the null hypothesis that the SMA coefficient is 0. The 95% confidence interval for the SMA coefficient is 0.3042$\pm$0.1058, which does not contain 0, so we reject the null hypothesis that the coefficient is 0. Thus the coefficient is significant.


Here are the forecasted values of the log strawberry prices.
```{r}
log_strawberry_forecast_sarima <- sarima.for(log(strawberry), n.ahead=8, p=0, d=1, q=0, P=0, D=0, Q=1, S=12)
log_strawberry_forecast_sarima
```

Here are the 95% Confidence Intervals of the log strawberry prices.

```{r}
matrix(c(log_strawberry_forecast_sarima$pred - qnorm(0.975)*log_strawberry_forecast_sarima$se,   log_strawberry_forecast_sarima$pred + qnorm(0.975)*log_strawberry_forecast_sarima$se), nrow = 8, ncol = 2)
```

```{r}
auto.arima(log(strawberry))
tsdiag(auto.arima(log(strawberry)))
```


```{r}
library(forecast)
auto.arima(strawberry)
forecast(auto.arima(strawberry), 14)
```


```{r}
window(log(strawberry), start = c(2022,10), end=c(2022,10))
```

Here are the forecasted values of the strawberry prices.

```{r}
exp(log_strawberry_forecast_sarima$pred)
```

Here are the confidence intervals for the forecasted strawberry prices.

```{r}
matrix(exp(c(log_strawberry_forecast_sarima$pred - qnorm(0.975)*log_strawberry_forecast_sarima$se,   log_strawberry_forecast_sarima$pred + qnorm(0.975)*log_strawberry_forecast_sarima$se)), nrow = (size(log_strawberry_forecast_sarima)/2), ncol = 2)
```



```{r}
acf(avocado)
pacf(avocado)
```

# Analysis and Forecasting of strawberry Prices using ARMA-GARCH Model